%!TEX root=paper.tex

\section{Introduction} 
% First introduce the general problem: data visualization on dashboards.
Data visualization on dashboards are widely used by companies and non-profit
organizations in various industries to inform decision making. Such
visualizations are often driven by complex data pipelines, where multiple data
sources are transformed. The transformations such as joins, filters, and
aggregations are applied to raw data. 
% introdce the metric after data visualization on dashboards.
Especially, metrics are commonly used to summarize and track key performance
indicators such as revenue, conversion rate, and customer lifetime value.
However, these visualizations typically present only the final aggregated
values. The intermediate computation steps, such as the underlying data sources,
transformation logic, and component metrics, are hidden. When the context or
derivation of a metric is unclear, users may misinterpret trends, struggle to
diagnose anomalies, or lose trust in the reported numbers~\cite{}. Therefore,
improving transparency into how metrics are computed, including their
intermediate components and data provenance, is critical for helping users
accurately interpret dashboard visualizations and make informed decisions.

\wc{@Elisa I changed the the general problem be data visualization on dashboards. I think only mentioning dashboards are not enough when we begin the paper.}

% Introduce what data lineage and provenance systems are, and how they can help
% with understanding the data transformation steps. Then, introduce the
% limitations of existing data lineage and provenance system.
Data lineage and provenance systems are designed to capture the flow of data
through various transformations in data pipelines usually visualized as graphs.
They have been developed to support data quality monitoring and governance of
data pipelines. Commercial software such as Collibra~\cite{} and
OpenMetadata~\cite{} capture dependencies between datasets, tables, and
attributes, enabling data professionals to trace the origin and flow of data.
While lineage graphs can reveal where a metric's underlying data originates, the
transformation logic often remains in SQL models or ETL scripts. As a result,
understanding how a metric is computed still requires inspecting source code,
which creates additional effort even for technical users. Moreover existing
lineage systems focus on dataset-level lineage rather than explaining the
semantics of transformations that define business metrics. This limitation is
more pronounced for non-technical stakeholders, who care less about which tables
or columns produce a metric but more about how to interpret the metric itself.
In practice, many analytic questions begin with a dashboard metric (e.g., ``Why
did revenue drop?'') rather than a dataset.
% 
We highlight the challenges of understanding how metrics manually are computed
through the following example.

\input{figures/example.tex}

% An example of the challenges of understanding how metrics are computed.
\begin{example}
Sheldon is a data analyst at an e-commere company. He is responsible for
monitoring the predictive liftetime value (PLTV) of customers, which can be a
great metric to understand the long-term value of custmers. He got granted
access to a dashboard that shows the PLTV metric, but he has no idea how the
PLTV metric is computed. He wants to understand how the PLTV metric is computed,
and what are the intermediate metrics that contribute to the final PLTV metric.
The transformation steps he sees on the dashboard are shown in the left of
Figure~\ref{tab:metrics}. From those transformation steps, he needs to go
through all the steps forward and backward to understand how the PLTV metric is
computed. First, in T100, he finds that three metrics, \texttt{prob\_active},
\texttt{expected\_orders}, and \texttt{expected\_order\_value}, are used to
compute the PLTV metric. Then, he needs to understand how those three metrics
are computed. And go through the same process for each of the rest of the
metrics until he reaches the raw data. 
\end{example}

This example highlights the challenges of understanding how metrics are
computed, especially when the transformation logic is complex and involves
multiple intermediate metrics. It requires users to manually explore the
transformation steps, and understand how those steps relate to the final metric.
This process is tedious and requires users to understand the technical details
of the transformation logic, which can be a barrier for non-technical users.
However, even though the dashboard can show the metric lineage and the
transformation, it still requires what are those metrics requiring the
explanations. For instance, to understand what the \texttt{prob\_active} metric
means, users need to understand the transformation steps that define the
\texttt{prob\_active} metric. Therefore, there is a need for a system that can
automatically extract and visualize the transformation steps that define a
metric, and provide explanations of what those steps mean in terms of the final
metric. We highlight the need for a system to explain the meaning of metrics by
showing the transformation steps that define the mtric in the follwing example.
\wc{Here, showing the transformation steps sounds quite confusing since we can just show the transformation steps without showing the meaning of those steps. I'll change the term soon.}

\begin{example}
After manual exploring, Sheldon extracts the metric lineage shown in the right
of Figure~\ref{tab:metrics}. From the metric lineage, he can see that the PLTV
metric is computed using various metrics such as \texttt{prob\_active},
\texttt{expected\_orders}, and \texttt{expected\_order\_value}. And also, each
of those metrics is computed using various sub-metrics. However, he still does
not what those metrics mean, specifically, what are the definitions of those
metrics, and how they are computed. For instance, to see how the
\texttt{prob\_active} metric is computed, he needs to see the transformation
steps more closely. He finds that the \texttt{prob\_active} metric is computed
using \wc{Elisa please put the transformation steps here.} To understand what
the metric means, he needs to go through all the transformation steps, and
understand what those steps mean.
\end{example}

Above example shows that even if the dashboard can show the metric lineage and
the transformation steps, it still requires users to understand what those
metrics mean, and how they are computed. To understand the meaning of a metric,
users need to understand the transformation steps that define the metric, and
how those steps relate to the final metric. This process is tedious and requires
users to understand the technical details of the transformation logic, which can
be a barrier for non-technical users. Therefore, there is a need for a system
that can automatically extract and visualize the transformation steps that
define a metric, and provide explanations of what those steps mean in terms of
the final metric. Such a system would help users understand how metrics are
computed, and improve their trust in the reported numbers.

We introduce \clue, a system that automatically extracts and visualizes the
transformation steps that define a metric, and provides explanations of what
those steps mean in terms of the final metric. \clue captures the transformation
logic that defines a metric, and visualizes the metric lineage and the
transformation steps in an interactive dashboard. Users can explore the metric
lineage and the transformation steps, and see how each step contributes to the
final metric. 
\wc{We should discuss how could we get the metrics. Is it pre-defined? Or we can automatically extract the metrics from the transformation steps?}

To support the trust of data-driven decision-making, provenance information must
move beyond data-centric lineage and provide accessible, metric-centric
explanations that help non-technical users understand how complex metrics are
derived. We highlight the need for a system that shows lineage and provenance of
metrics.

\smallskip \noindent \emph{\underline{Related Work.}} 
\wc{@Elisa could you check the related work section if it looks good?}
% provenance graphs or relevant works
Provenance/lineage graphs~\cite{DBLP:journals/tods/ChapmanLMT24,
DBLP:journals/tvcg/CaiGXYWXWLW25,DBLP:conf/deem/PinaCK0M24} show the data flow
and transformations in various domains such as
databases~\cite{DBLP:journals/corr/abs-0912-5340, DBLP:conf/icdt/BunemanKT01},
and machine learning pipelines~\cite{DBLP:conf/deem/PinaCK0M24,
DBLP:journals/tods/ChapmanLMT24}. However, while they often focus on changes in
the data and transformations, they do not discuss what these transformations
mean in terms of metrics, and how the transformation meanings are related to the
final results. 
% visualizing data transformation
Visualization of intermediate
transformations~\cite{DBLP:journals/pvldb/LeeTABCKMSYHP21,
DBLP:conf/chi/WangLDM025} allow users to transform and explore data
interactively. DataFormulator2~\cite{DBLP:conf/chi/WangLDM025} provides an
environment for transforming and visualizing intermediate results, while
Lux~\cite{DBLP:journals/pvldb/LeeTABCKMSYHP21} surfaces intermediate insights
directly in Jupyter notebooks. However, neither discusses how transformations
relate to the final results in terms of metrics.

\smallskip \noindent \emph{\underline{Demonstration.}} In our demonstration,
participants will observe \clue's capability to X