%!TEX root=paper.tex

\section{Introduction} 
Data visualizations on dashboards are ubiquitous across industries and play a
central role in supporting decision-making at all organizational levels. Such
visualizations are often driven by complex data pipelines, where data from
multiple sources is integrated and transformed using operations such as joins,
filters, and aggregations. Through a series of transformations, raw data is
converted into business metrics on dashboards, enabling organizations to
summarize and track key performance indicators such as revenue, conversion rate,
and customer lifetime value. These metrics are essential for monitoring business
health, identifying trends, and making informed decisions. However, the
intermediate transformation steps and underlying data--source data and
transformation logic--are often hidden, as these visualizations typically
present only the final aggregated metrics. This lack of transparency into how
metrics are computed leads to challenges for users unfamiliar with them, causing them to misinterpret
trends, struggle to diagnose anomalies, or lose trust in the reported
numbers~\cite{matheus2020data}.

% Introduce what data lineage and provenance systems are, and how they can help
% with understanding the data transformation steps. Then, introduce the
% limitations of existing data lineage and provenance system.
 A common way users engage with dashboard metrics is through goal-driven
 \texttt{how-to} questions such as ``\textit{How can we increase this
 metric?}''. However, without a clear understanding of how the metric is
 calculated, these questions become difficult to answer. Data lineage and
 provenance systems~\cite{galhotra2022hyper, DBLP:conf/sigmod/MeliouS12} are
 designed to expose this information by capturing the flow of data through
 transformations in data pipelines. Commercial software such as
Commercial software such as Collibra~\cite{collibra}
 % and OpenMetadata~\cite{openmetadata} 
 captures dependencies between databases, tables, and attributes, enabling both
 technical and non-technical users to trace the origin and flow of data. While
 these lineage graphs can reveal where a metric's underlying data originates,
 the transformation logic often remains in SQL models or ETL scripts. As a
 result, understanding how a metric is computed still requires inspecting source
 code, which creates additional effort even for data professionals. Moreover,
 existing lineage systems focus on dataset-level lineage rather than explaining
 the semantics of transformations that define business metrics. This limitation
 is more pronounced for non-technical stakeholders, who care less about which
 tables or columns produce a metric, but more about how to interpret the metric
 itself. 
%
In summary, these limitations mean current systems cannot answer \texttt{how-to}
questions since (1)~they do not extract and represent metric lineage
automatically to show how metrics are related, and (2)~they lack interactive
explanations of how such metrics can be increased or decreased.
% 
We first highlight the challenges of manually understanding metric lineage
through the following example.

\input{figures/example.tex}

\begin{example}
Sheldon is a new-hired data analyst at an e-commerce company and recently assigned a
project to grow the predictive lifetime value (\metric{PLTV}) of customers. \metric{PLTV} is a key business metric that estimates the total revenue a customer is expected to generate over their relationship with the company, making it central to decisions around customer acquisition, retention, and long-term growth. As someone new to the company’s operations, Sheldon is not yet familiar with what drives \metric{PLTV}. Although he is given access to a dashboard that reports the metric, it doesn't provide explanation of how \metric{PLTV} is computed. To answer his question ``How can we increase \metric{PLTV}?'', Sheldon must first understand how it is defined.
%
As the logic is not explicitly shown in the dashboard and the company's data lineage system, Sheldon
needs to look at the underlying implementation. He manually inspects the SQL code that defines the metric, and examines the
sequence of transformation steps shown in Figure~\ref{tab:metrics}. From these
steps, he attempts to trace how \metric{PLTV} is derived. He first observes that
in T100, \metric{PLTV} is computed using three intermediate metrics:
\metric{prob\_active}, \metric{expected\_orders}, and
\metric{expected\_order\_value}. He then recursively traces each of these
metrics backward through the transformation pipeline to understand their
origins. 

When tracing \metric{expected\_order\_value}, Sheldon discovers that it depends
on a sequence of transformations (T20, T21, T22, T50, T97, T98, T99). These
transformations start from raw data and are joined and aggregated to compute
intermediate metrics. T50 derives the \metric{\#\_orders} submetric, while (T20,
T21, T22, T97, T98) derive \metric{\#\_total\_gross\_value}, which are then used
to compute \metric{expected\_order\_value} in T99. Sheldon now needs to again
manually inspect for tracing \metric{prob\_active} and \metric{expected\_orders}
to fully understand how \metric{PLTV} is computed. %\ew{Do you think the number here should align with what we show in demo? If so, then we only have 3 submetrics and each has two child nodes.} \wc{Yes, but for now I made the examples as a approximate 3-ary tree though.} 
Sheldon finds that there are
5 submetrics arranged as an approximate 3-ary tree of depth 5, giving at most
$\frac{3^5-1}{2} = 121$ nodes, $120$ directed edges, and $3^4 = 81$ root-to-leaf
paths, requiring at most $120 + 81 = 201$ edges and paths, hence $2 \times 201 =
402$ traversals forward and backward. Since Sheldon must hold the entire lineage
structure, this process makes him lose track of the metric lineage of
\metric{expected\_order\_value} after just a few steps through \metric{prob\_active}.
\end{example}

% \wc{We need to decide if we're going to use sub-metric or submetric.} \ew{Submetric is commonly used in the context of BI tool.}
This example highlights the challenges of understanding how metrics are
computed, especially when the transformation logic is complex and involves
multiple intermediate metrics. It requires users to manually explore the
transformation steps and understand how those steps relate to the final metric.
This process can extract the metric lineage, showing how the final metric is
derived from intermediate metrics, and how those intermediate metrics are
derived from other submetrics. Tools such as SQLGlot~\cite{sqlglot} can help
users understand transformation steps by parsing SQL code. However, SQL parsing
alone does not help users understand how metrics are related. Therefore, a
system is needed to automatically extract metric lineage and represent
relationships among metrics.

In addition to extracting metric lineage to answer \texttt{how-to} questions,
users must still interpret what each intermediate metric \textit{means}, not
just how it connects. For example, even if Sheldon can see that \metric{PLTV}
depends on \metric{expected\_order\_value}, the structural relationship alone is
not enough, he still needs to understand how to increase \metric{PLTV} by
adjusting \metric{expected\_order\_value} and other submetrics. However, since
there are multiple submetrics and multiple adjustment strategies, Sheldon cannot
easily determine how changes in individual intermediate submetrics would affect
\metric{PLTV}. This points to the need for a system that can suggest feasible
adjustment strategies over dependent metrics and propagate those changes to
satisfy a target goal. We highlight the need for explaining the meaning of
transformations in the following example.

\begin{example}
After manually exploring the transformation pipeline, Sheldon extracts the
metric lineage shown in the right of Figure~\ref{tab:metrics}. From the
\metric{PLTV} lineage, he can see that \metric{PLTV} is computed from multiple
intermediate metrics, including \metric{prob\_active},
\metric{expected\_orders}, and \metric{expected\_order\_value}, each of which in
turn is derived from other submetrics.
%
Despite understanding these dependencies, Sheldon encounters new challenges when
trying to answer ``How can we change \metric{PLTV} value?'' Specifically, he
wants to know ``How can we increase \metric{PLTV} by 20\%?'' and ``How to
achieve a certain increase in \metric{PLTV}?''.  Multiple adjustment strategies
are possible, for example, he could choose to increase one of the three
intermediate metrics by 20\%, or distribute the increase across multiple metrics
, such as 10\% for \metric{prob\_active} and 10\% for \metric{expected\_orders}.
However, Sheldon cannot determine how changes in the intermediate metrics would
propagate to the final metric, and how to adjust the intermediate metrics to
achieve the desired change in \metric{PLTV}. 
%
With up to 5 levels of metrics, each with 3 submetrics, there are up to $2^{120}
\approx 10^{36}$ possible strategies across all 120 nodes. Moreover, determining
the adjustment range for each metric is itself a non-trivial task, further
increasing the search space. With such a large number of possibilities, Sheldon
cannot easily determine how changes in individual intermediate submetrics would
affect \metric{PLTV}. Even if he could compute the expected change in
\metric{PLTV} for each strategy, he would still need to manually compare them
all to determine which is best.
\end{example}

The example above illustrates the challenges of answering \texttt{how-to}
provenance questions on metric lineages. Even when a user understands the metric
lineage, determining which metrics to adjust and by how much to achieve a
desired outcome requires navigating an exponentially large search space. This
motivates the need for a systematic approach that can automatically identify
effective adjustment strategies and explain how changes in component metrics
propagate to the final metric.

We introduce \clue, a system that automatically extracts and visualizes the
transformation steps that define a metric, and provides simulation capabilities
to support metric adjustment. \clue captures the transformation logic that
defines a metric, and visualizes the metric lineage and the transformation steps
in an interactive dashboard. Users can explore the metric lineage and the
transformation steps, and see how each step contributes to the final metric. As
users explore the transformation steps, they can also see how changes in the
final metric change intermediate metrics, and how to adjust intermediate
metrics and data. \clue is built upon a SQL-based data pipeline where metrics
are defined through multi-layer transformations. Assuming that dashboard
metadata and the upstream data pipeline are accessible, CLUE automatically
extracts transformation operations and metric dependencies from those sources.

To support the trust of data-driven decision-making, provenance information must
move beyond data-centric lineage and provide accessible, metric-centric
explanations that help non-technical users understand how complex metrics are
derived. We highlight the need for a system that shows the lineage and provenance of
metrics.

\smallskip \noindent \emph{\underline{Related Work.}} 
% \wc{@Elisa could you check the related work section if it looks good?}
% provenance graphs or relevant works
Provenance/lineage graphs show the data flow and transformations in various
domains such as databases~\cite{DBLP:conf/sigmod/MeliouS12,galhotra2022hyper}
for \texttt{how-to} questions, and machine learning
pipelines~\cite{DBLP:conf/deem/PinaCK0M24, DBLP:journals/tods/ChapmanLMT24}.
However, while they often focus on changes in the data and transformations, they
do not discuss what these transformations mean in terms of metrics, and how the
transformation meanings are related to the final results. 
% visualizing data transformation
Visualization of intermediate
transformations~\cite{DBLP:journals/pvldb/LeeTABCKMSYHP21,
DBLP:conf/chi/WangLDM025} allow users to transform and explore data
interactively. DataFormulator2~\cite{DBLP:conf/chi/WangLDM025} provides an
environment for transforming and visualizing intermediate results, while
Lux~\cite{DBLP:journals/pvldb/LeeTABCKMSYHP21} surfaces intermediate insights
directly in Jupyter notebooks. However, neither discusses how transformations
relate to the final results in terms of metrics.

\smallskip \noindent \emph{\underline{Demonstration.}} In our demonstration,
participants will observe \clue's capability to decompose complex metrics into
interpretable components, reveal how they are computed from underlying data, and
support goal-driven exploration by simulating actionable changes across the
metric hierarchy. We provide an overview of CLUE in Section 2 and a detailed
walkthrough of our demonstration scenario in Section 3. 